The imprint of global magma oceans on exoplanet demographics

### 2023-04-03 src/tex/ms.tex
remove bullets
edit optimistic case and PLATO

% Main body
\section{Introduction}
%\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021}.}
%\todo[inline]{``Points for the intro:
%\bioverse\\
%Transit survey to measure mass/radius of planets\\
%Does the HZ have an inner edge?\\
%Run-away GH\\
%Magma oceans\\
%Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman. Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, LIFE.\\
%~\\
%In contrast, PLATO and \nautilus\ are \\
%Possibility of follow-up to measureme masses
%''}

\begin{itemize}
    \item hot/warm planets are best studied
    \item runaway greenhouse climate (Ingersoll1969,Kasting1988,Nakajima1992). Transition is sharp in instellation.
    \item HZ concept.''(Kasting et al. 1993; Kopparapu et al. 2013, 2014)'' Relevance in the search for habitable planets and life. (Science gap:) Does it exist? Can it be tested? => statistically detect inner edge
    \item Bioverse
    \item Transit survey => PLATO
    \item Using PLATO as a baseline survey, we do A, B, C.
\end{itemize}

\begin{note}
Recent years have brought tremendous advances in our understanding of terrestrial-sized planets, mainly due to space-based transit surveys such as \kepler\ \citep{Borucki2010} and \tess\ \citep{Ricker2014a} that pushed detection sensitivities into the terrestrial regime.
Due to the inherent detection biases of the transit technique, most current well-probed small planets are much closer to their host star than Earth is to the Sun. CITE
These bodies experience climatic conditions that are similar to the environment of the inner Solar System planets at early stages of their formation. CITE
Studies of the geophysical state and evolution of warm exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds.

A much regarded object of study are the divergent atmospheric evolutions of Earth and Venus.
While both have accreted from a similar mass reservoir~\citep{Morbidelli2012} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{Kasting1988,Hamano2013,Kane2014,Turbet2021}. "(Donahue et al. 1982; Bézard, B. \& de Bergh, 2007; Way et al. 2016; Way \& Del Genio 2020; )."
Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.
%magma oceans...
However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet.
Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits outgoing radiation and thus the release of excess heat~\citep{Ingersoll1969,Kasting1988,Wordsworth2022}.
This runaway greenhouse state can not only extend the magma ocean stage of a rocky planet to hundreds of \SI{}{\mega\year}, it can also lead to a loss of its entire water reservoir by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen. CITE
Venus' atmospheric composition is consistent with the planet having gone through this phase, although the past presence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016}.
A water-dominated atmosphere of a planet in the runaway greenhouse state is significantly inflated~\citep{Turbet2019,Mousis2020}, increasing its transit radius by an amount mainly depending on the surface water content.
An Earth-like planet experiences a \SI{\sim 500}{\kilo\meter} radius increase upon crossing the threshold instellation, and its transit depth increases by \SI{10}{\ppm} to \SI{1000}{\ppm} depending on the spectral type of the host star~\citep{Turbet2019}.
It was further found that the redistribution of water between the steam atmosphere and the molten mantle leads to additional, non-negligible deviations in atmospheric scale-heights and transit radii~\citep{Dorn2021}.

Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{Hamano2013}. CITE
Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone, whose prevalent definition refers to the possibility of liquid water on an Earth-like planet with a CO$_2$/H$_2$O/N$_2$ atmosphere~(\citealp[][but other definitions have been put forward, e.g., ]{Kasting1993,Kopparapu2013,Kopparapu2014} \citealp{Hall2023,Spinelli2023}).
The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1853,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in NASA's Astrophysics Data System returned $\sim\num{1700}$ results.}.
Investigating its boundaries in parameter space has helped improve our understanding of habitability. CITE
Nevertheless, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017}.
The question naturally emerges if a habitable zone -- in its common form as boundaries in receiving instellation -- indeed exists and how the habitable zone hypothesis can be tested.
Observational tests being considered include searches for evidence of liquid water conveyed by ocean glint~\citep{Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019}.
Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos ~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
All these tests require surveying a large enough sample of terrestrial-sized planets with next-generation instruments, rendering them out of reach in the immediate future.

Here, we explore the feasibility of a statistical test of the habitable zone hypothesis by surveying planetary bulk properties close to its inner edge, the runaway greenhouse transition.
Our goal is to assess the ability of near-future transit surveys to test the hypothesis that the runaway greenhouse effect causes an abrupt change of planetary radii and bulk densities when ordered by receiving instellation~\citep{Turbet2019}.
Our main tool for this is \bioverse, a simulation framework for assessing the statistical power of exoplanet surveys~\citep{Bixel2021}.
It consists of a sample generator that populates stars from the \gaia\ catalog~\citep{Gaia_mission,Gaia_DR3}  with planetary systems based on state-of-the-art occurrence rates, a flexible survey simulator that allows for a broad range of trade studies, and a hypothesis testing module that quantifies the survey's ability to detect a previously injected trend.
Trying different instrumentation and survey designs, we utilize \bioverse\ to recover runaway greenhouse-induced effects based on model predictions~\citep{Turbet2020,Dorn2021} that we inject into a baseline planet population.

In particular, we test the capability of the \plato\ \citep[PLAnetary Transits and Oscillation of stars, ][]{Rauer2016} mission, which will measure the radii of a large number of terrestrial-sized planets, to detect the runaway greenhouse transition and identify the astrophysical unknowns on which this depends.
We then perform a parameter study to explore which trade-offs in the survey design of a \plato-like mission optimize the chance to detect and characterize the runaway greenhouse transition.

We organize the paper as follows:
Section~\ref{sec:met_baseline} introduces our baseline model produce synthetic star and planet samples.
    In Sect.~\ref{sec:met_rghmodel}, we describe the model component that produces runaway greenhouse-induced radius changes.
    Section~\ref{sec:met_surveys-hypotests} explains our survey simulations and hypothesis tests.
    We present our results in Sect.~\ref{sec:results} before interpreting them in Sect.~\ref{sec:discussion}.
    Finally, we summarize our findings in Sect.~\ref{sec:conclusions}.
\end{note}

@@

For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated radius due to a steam atmosphere.
While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Abe1988,Kasting1988,Nakajima1992,Goldblatt2013,Leconte2013,Boukrouche2021}.
To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
Their calculations rely on the same mass-radius relations for rocky interiors that we apply for our non-runaway planets~\citep{Zeng2016}.
For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
This scenario does not take into account the expected additional radius change due to molten interiors~\citep{Bower2019}.
It further assumes a \textit{dry melt} without dissolved volatiles.

@@

\section{Results}\label{sec:results}
\subsection{Statistical signature of the runaway greenhouse threshold}\label{sec:res_signature}
To characterize the population-level imprint of individual radius changes, we generated a generic planet population with an injected runaway greenhouse effect assuming a water fraction $x_\mathrm{H_2O} = \var{wrr}$.
\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
        \caption{Synthetic planets above and below the runaway greenhouse threshold.
        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of net instellation. Only the planets marked as transiting are observable.
            Above the runaway greenhouse threshold $S_\mathrm{thresh} = \var{injectedSthresh}$, some planets maintain their original radii (green crosses) whereas some have their radii inflated (blue crosses) by the amount indicated with gray lines.
        The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue line).
            This runaway greenhouse hypothesis can be tested against the null hypothesis $H_0$ (green line), where average radii are independent of instellation.}
        \label{fig:HnullHmo}
    \end{centering}
\end{figure}
Figure~\ref{fig:HnullHmo} shows the resulting planetary radii.
When ordered in orbital period space, the different planet types overlap, diluting the demographic imprint.
With net instellation as an independent variable, planets above and below the runaway greenhouse threshold separate: The runaway greenhouse-induced radius inflation introduces a discontinuity of average planet radii and bulk densities as a function of stellar irradiation.
We also show the predictions of observable average planet radii from the statistical hypotheses defined above.
%The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.
Within the runaway greenhouse regime, an average radius change of \avgRadiusChange\ occurs.
This pattern is consistent with the injected radius inflation as predicted from the atmospheric models (see Appendix~\ref{app:model-pop_comparison} for an investigation of the interplay between model predictions and our synthetic planet population).

@@

\subsubsection{OPTIONAL...: Atmospheric signatures}\label{sec:dis_atmospheres}
\todo[inline]{TIM: Ideas?\\
discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g. Tim's recent talk). Or: high O2 abundance during H loss (Wordsworth \& Pierrehumbert 2013b; Luger \& Barnes 2015, Schaefer et al. 2016)}
\todo[inline]{``Can a smaller sample size's diagnostic value be increased if atmospheric compositional information can be added, i.e., PLATO + ARIEL (or similar combo)?''}
\todo[inline]{also: use atmospheric characterization to identify small planet cores with thick H/He envelopes that contribute to diluting the signal (aka background population). These will be amenable for transmission spec due to large scale heights (Moran2018, \citet{Turbet2019}).\\ Spectral water features  $<<$ rgh-induced radius change~\citep[][end of Sect. 4]{Turbet2019}.}

@@

Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity~\citep{Turbet2021,Pierrehumbert2022}.
The choice of statistical estimator for the hypothesis tests may further blur the recovered discontinuity.
We compare our nominal running mean approach with a binned statistic in Appendix~\ref{app:binnedstats}.

@@

\section{Conclusions}\label{sec:conclusions}
Runaway greenhouse climates on rocky planets cause significant inflation of their radii.
Using \bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius discontinuity resulting from this inflation in the exoplanet population.
Our key findings are as follows:
\begin{enumerate}
    \item The predicted runaway greenhouse transition causes a discontinuity in the radius and density distribution of small exoplanets with respect to the incident host star irradiation.
    \item The demographic imprint of the runaway greenhouse threshold should be detectable with high-precision transit measurements. For a large sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of the observed planets and if typical water mass fractions are above $\sim 10^{-3}$.
    \item We find that the planned \plato\ transit survey will provide a sufficient sample and the required precision to detect the predicted trend. Assuming the projected photometric precision, \plato\ will be able to test the runaway greenhouse hypothesis for planet yields $\gtrsim 100$.
    \item The diagnostic power of transit missions in testing this hypothesis can be increased through a follow-up campaign providing planet mass measurements.
          This can reduce the required planet yield by about a factor of three.
          Only an adequate sample of planets orbiting M~dwarfs will ensure sufficient targets on both sides of the expected threshold instellation.
    \item Testing the runaway greenhouse hypothesis on a population level can provide constraints on the water inventory of rocky exoplanets and thus make an important contribution to assessing their habitability.
         A detection will provide an empirical confirmation of the habitable zone concept and localize its inner edge.

@@

%\begin{acknowledgments}
%The authors thank Gabriele Cugno, Yann Fabel, Paul Molliere, Gijs Mulders, Matthew Murphy, and Terry-Ann Suer for insightful discussions.
%We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.
%This work has made use of data from the European Space Agency (ESA) mission {\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia} Data Processing and Analysis Consortium (DPAC, \url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC has been provided by national institutions, in particular the institutions participating in the {\it Gaia} Multilateral Agreement.
%\end{acknowledgments}

>>>>

% Main body
\section{Introduction}
\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
\todo[inline]{``Points for the intro:
\bioverse\\
Transit survey to measure mass/radius of planets\\
Does the HZ have an inner edge?\\
Run-away GH\\
Magma oceans\\
Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman. Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, LIFE.\\
~\\
In contrast, PLATO and \nautilus\ are \\
Possibility of follow-up to measureme masses
''}

\todo[inline]{introduce the HZ Inner Edge Discontinuity\textsuperscript{\texttrademark}}

\begin{note}
For observations of rocky exoplanets, the currently best-probed regime is that of warm, close-in planets.
These bodies experience climatic conditions that are similar to the environment of the inner Solar System bodies at early stages of their formation.
Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds.

...
    Venus and Earth, while having accreted from the same mass reservoir and despite their similar bulk properties, evolved into planets with very different environmental conditions on their surfaces.
    At the formation of the Solar System, both planets underwent a giant impact phase [CITE!] that melted their mantles, leading to a magma ocean stage.
    Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.
    \citet{Hamano2013} showed that such a late desiccation process is not needed to explain the stark differences: for a planet rich in volatiles and receiving high enough radiation levels from its host star, a steam atmosphere limits the outgoing radiation from the (molten) planetary surface.
    This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}, enough to remove the entire water reservoir from a rocky planet by hydrodynamic escape.
    \todo[inline]{important for motivating the existence of the pattern: ``the solidification timescale can become comparable to the main-sequence lifetime of the star (Hamano et al. 2013, 2015)'' \citep{Lichtenberg2022} }
    Consequently, it is aptly considered to be the inner boundary of the habitable zone~\citep{Kasting1993,Kopparapu2013}.

    (Under the right conditions, different orbital distances alone can thus decide about a planet's water content and ultimately its habitability.)
    If received flux makes the difference between our own habitable planet and dry, dead Venus, there is good reason to believe that such patterns exist in exoplanet systems as well.

    ...long-term goal/overarching objective: derive the geophysical history of rocky extrasolar planets. More concrete: Constrain the limits of runaway greenhouse transitions, and thus the inner edges of the habitable zone. Also: How close is Earth from this runaway greenhouse irradiation limit?
   ...
\todo[inline]{introduce magma oceans and their influence on planetary radii~\citep{Dorn2021}.}
    ...current/future observations of planets that are currently in the runaway greenhouse phase may constrain properties of their planetary mantles... make connection between interior and atmospheres.
    ...

    The runaway greenhouse transition is a robust prediction from climate models, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
On the planetary level, however, such undertaking is hindered by degeneracies: From internal composition to atmospheric loss to volatile content, many factors influence a planet's size.
It is therefore prudent to shift the focus to the planet population as a whole and search for a statistical effect in exoplanet demographics~\citep{Turbet2019}.
 Using \bioverse\ to combine the predictions from geophysical models of steam atmospheres with a synthesis of planets informed by observed demographics, we...

    Planetary radius changes due to the combined effect of steam atmospheres and water dissolved in the molten mantle of planets within the runaway greenhouse threshold is expected to cause a discontinuity in the radius distribution of small (DEFINITION) exoplanets.

\end{note}

@@

For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated radius due to a steam atmosphere.
While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Abe1988,Kasting1988,Nakajima1992,Goldblatt2013,Leconte2013,Boukrouche2021}.
To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
Their calculations rely on the same mass-radius relations for rocky interiors that we apply for our non-runaway planets~\citep{Zeng2016}.
For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
This scenario does not take into account the expected additional radius change due to molten interiors~\citep{Bower2019}.
It further assumes a \textit{dry melt} without dissolved volatiles.

@@

\section{Results}
\subsection{Statistical signature of the runaway greenhouse threshold}\label{sec:res_signature}
To characterize the population-level imprint of individual radius changes, we generated a generic planet population with an injected runaway greenhouse effect assuming a water fraction $x_\mathrm{H_2O} = \var{wrr}$.
\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
        \caption{Synthetic planets above and below the runaway greenhouse threshold.
        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of net instellation. Only the planets marked as transiting are observable.
            Above the runaway greenhouse threshold $S_\mathrm{thresh} = \var{injectedSthresh}$, some planets maintain their original radii (green crosses) whereas some have their radii inflated (blue crosses) by the amount indicated with gray lines.
        The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue line).
            This runaway greenhouse hypothesis can be tested against the null hypothesis $H_0$ (green line), where average radii are independent of instellation.}
        \label{fig:HnullHmo}
    \end{centering}
\end{figure}
Figure~\ref{fig:HnullHmo} shows the resulting planetary radii.
When ordered in orbital period space, the different planet types overlap, diluting the demographic imprint.
With net instellation as an independent variable, planets above and below the runaway greenhouse threshold separate: The runaway greenhouse-induced radius inflation introduces a discontinuity of average planet radii and bulk densities as a function of stellar irradiation.
We also show the predictions of observable average planet radii from the statistical hypotheses defined above.
%The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.
Within the runaway greenhouse regime, an average radius change of \avgRadiusChange\ occurs.
This pattern is consistent with the injected radius inflation as predicted from the atmospheric models (see Appendix~\ref{app:model-pop_comparison} for an investigation of the interplay between model predictions and our synthetic planet population).

@@

\subsubsection{OPTIONAL...: Atmospheric signatures}\label{sec:dis_atmospheres}
\todo[inline]{TIM: Ideas?\\
discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g. Tim's recent talk)}
\todo[inline]{``Can a smaller sample size's diagnostic value be increased if atmospheric compositional information can be added, i.e., PLATO + ARIEL (or similar combo)?''}
\todo[inline]{also: use atmospheric characterization to identify small planet cores with thick H/He envelopes that contribute to diluting the signal (aka background population). These will be amenable for transmission spec due to large scale heights (Moran2018, \citet{Turbet2019}).\\ Spectral water features  $<<$ rgh-induced radius change~\citep[][end of Sect. 4]{Turbet2019}.}

@@

Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, or surface gravity~\citep{Pierrehumbert2022}.
The choice of statistical estimator for the hypothesis tests may further blur the recovered discontinuity.
We compare our nominal running mean approach with a binned statistic in Appendix~\ref{app:binnedstats}.

@@

\section{Conclusions}
Runaway greenhouse climates on rocky planets cause significant inflation of their radii.
Using \bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius discontinuity resulting from this inflation in the exoplanet population.
Our key findings are as follows:
\begin{enumerate}
    \item The predicted runaway greenhouse transition causes a discontinuity in the radius and density distribution of small exoplanets with respect to the incident host star irradiation.
    \item The demographic imprint of the runaway greenhouse threshold should be detectable with high-precision transit measurements. For a large sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of the observed planets and if typical water mass fractions are above $\sim 10^{-3}$.
    \item We find that the planned \plato\ transit survey will provide a sufficient sample and the required precision to detect the predicted trend. Assuming the projected photometric precision, \plato\ will be able to test the runaway greenhouse hypothesis for planet yields $\gtrsim 100$.
    \item The diagnostic power of transit missions in testing this hypothesis can be increased through a follow-up campaign providing planet mass measurements.
          This can reduce the required planet yield by about a factor of three.
          Only an adequate sample of planets orbiting M~dwarfs will ensure sufficient targets on both sides of the expected threshold instellation.
    \item Testing the runaway greenhouse hypothesis on a population level can provide constraints on the water inventory of rocky exoplanets and thus make an important contribution to assessing their habitability.
         A detection will provide an empirical confirmation of the habitable zone concept and localize its inner edge.

@@

%\begin{acknowledgments}
%The authors thank Gabriele Cugno, Yann Fabel, Paul Molliere, Gijs Mulders, Matthew Murphy, and Terry-Ann Suer for insightful discussions.
%We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.
%\end{acknowledgments}
